---
title: "Chapter 1 · My EDA project"
author: "Florian Escaffre"
date: today
---

> **How to fill this chapter in**
> This is your main portfolio piece. The numbered sections below follow the
> **IMRAD** structure of research: question ≈ *introduction*, getting the
> data + cleaning decisions ≈ *methods*, plotting the story ≈ *results*, and
> findings ≈ *discussion*. In the *Reproducible Workflows* course:
>
> - **Session 2** lab: find a dataset you care about on an open-data portal
>   (data.gouv.fr or your own country's), source it **reproducibly by code**,
>   run an EDA, clean the data, and commit it to git.
> - **Session 3** work block: finish and polish the chapter, pin your data and
>   environment, open a pull request for review, and publish the book to a URL.
>
> Everything between the `TODO` markers below is a template to replace with
> your own content.

## The question

<!-- TODO: In a sentence or two, what question are you trying to answer with this dataset?  -->

## Getting the data

The data is sourced **by code** from an open-data portal of your choice. Because
the download lives inside the document, it is recorded in the analysis itself;
anyone who re-renders this document gets the same data and the same results.
Record the dataset's **publisher and URL** in the text so the provenance is
explicit.

The only portal-specific part is the *fetch*. This template shows two ways;
use the one that fits your portal:

- **`{rdatagouv}`** (data.gouv.fr only, optional): search/download from R.
- **Any portal by URL**: `download.file()` works with any portal's direct
  download link.

```{r}
#| label: setup
#| message: false
# Load what you use. {rdatagouv} only if you're on data.gouv.fr and want it.
library(rdatagouv)    # search / download datasets from data.gouv.fr
library(janitor)        # clean column names, tabulate with tabyl()
library(naniar)         # visualise missing values
library(skimr)
library(dplyr) 
library(tidyverse)       # one-line data summary with skim()
```

```{r}
#| label: download
#| message: false
board <- pins::board_folder("data/", versioned = TRUE)
ds <- pins::pin_read(board, "eco2mix_regional")
```

## A first look

```{r}
#| label: skim
#| fig-cap: "A one-line summary of the dataset."
# The skim() output gives you type, missingness and distributions at a glance.
ds |> 
  skimr::skim() 
```

```{r}
#| label: missingness
#| fig-cap: "Missing values across variables."
vis_miss(ds, "warn_large_data" = FALSE) #— where is data missing, and why might that be?

# datset too big to use vis_miss
```

## Cleaning decisions {#sec-cleaning}

<!-- TODO: document the decisions you made while cleaning the data. What did you
rename, recode, or drop, and why? A defensible cleaning decision is a feature
of a good analysis, not an afterthought. -->

```{r}
#| label: justif-drop-na-consommation

ds |> 
  janitor::clean_names() |>
  filter(is.na(consommation_mw)) |> 
  nrow()

ds |> 
  janitor::clean_names() |>
  mutate(rang = row_number()) |> 
  filter(is.na(consommation_mw)) |> 
  pull(rang)
```
```{r}
#| label: justif-drop-stockage-batterie

ds |> 
  janitor::clean_names() |> 
  summarise(
    valeurs_uniques_stockage = list(unique(stockage_batterie)),
    valeurs_uniques_destockage = list(unique(destockage_batterie))
  )
```

Vérifier les colonnes relatives au stockage des baterries : seulement des NA et 0
Décalage de 1h entre les colonnes date, heure et date_heure + missing values in date_heure : decision to trop the column date_heure
Erreur de détection de types des colonnes (str au lieu de double par exemple)
Les 12 premières lignes sont les seules avec des NA sur la consommation en MW, et des NA sur toutes les colonnes : dropped

```{r}
# Clean names, drop empty column, fix column type
ds_clean <- ds |> 
  janitor::clean_names() |>
  janitor::remove_empty(which = "cols") |> 
  select(-c("stockage_batterie", "destockage_batterie")) |> 
  mutate(
    heure = hms::as_hms(heure),
    date = as.Date(date),
    date_heure = lubridate::ymd_hms(date_heure, tz = "UTC"),
    code_insee_region = as.integer(code_insee_region),
    across(
      .cols = -c(region, nature, date, heure, date_heure, code_insee_region),
      .fns = as.double
    )
  ) |> 
  tidyr::drop_na("consommation_mw")
```

```{r}
#| label: justif-fuseaux-date-heure

lubridate::tz(ds_clean$date_heure)

ds_clean |> 
  mutate(
    heure_num = as.numeric(hms::as_hms(heure)) / 3600,
    heure_utc_num = lubridate::hour(date_heure) + lubridate::minute(date_heure)/60,
    diff = (heure_num - heure_utc_num) %% 24
  ) |> 
  count(diff)
```
```{r}
ds_clean |> 
  filter(is.na(date_heure) | is.na(heure)) |> 
  count(date) |> 
  arrange(desc(n))
```

```{r}
ds_clean |> 
  filter(is.na(date_heure)) |> 
  count(heure, region)
```
Missing values of date_heure of the hours corresponding to hour change in autumn and summer of each year : drop?
More than 50% of missing values on the tco and tch columns : need to drop ? / can they be useful ?


```{r}
#| label: skim2
#| fig-cap: "A one-line summary of the cleaned dataset."
# The skim() output gives you type, missingness and distributions at a glance.
ds_clean |> 
  skimr::skim() 
```



## Plotting the story

<!-- TODO: replace with your figures. Give every figure a caption and refer to
it in the text (e.g. "Fig. 1 shows that …"). If a figure uses randomness
(sampling, jitter, a simulation), fix the seed *inside* that cell with
`withr::with_seed(2025, { ... })` so the figure is identical on every render. -->

### Energy mix by region
```{r}
#| label: mix-par-region

mix_region <- ds_clean |> 
  select(region, thermique_mw, nucleaire_mw, eolien_mw, solaire_mw, 
         hydraulique_mw, bioenergies_mw) |> 
  pivot_longer(-region, names_to = "filiere", values_to = "mw") |> 
  group_by(region, filiere) |> 
  summarise(prod_totale = sum(mw, na.rm = TRUE), .groups = "drop_last") |> 
  mutate(part_pct = prod_totale / sum(prod_totale) * 100) |> 
  ungroup()

mix_region
```

```{r}
#| label: viz-mix-par-region

couleurs_filieres <- c(
  "nucleaire_mw" = "#F4A300",     # orange (convention RTE/IEA courante)
  "thermique_mw" = "#8B0000",     # rouge foncé (fossile)
  "eolien_mw" = "#87CEEB",        # bleu ciel
  "solaire_mw" = "#FFD700",       # jaune (soleil)
  "hydraulique_mw" = "#0047AB",   # bleu foncé (eau)
  "bioenergies_mw" = "#556B2F"    # vert olive
)

mix_region |> 
  ggplot(aes(x = region, y = part_pct, fill = filiere)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = couleurs_filieres,
    labels = c(
      "nucleaire_mw" = "Nucléaire",
      "thermique_mw" = "Thermique (fossile)",
      "eolien_mw" = "Éolien",
      "solaire_mw" = "Solaire",
      "hydraulique_mw" = "Hydraulique",
      "bioenergies_mw" = "Bioénergies"
    )
  ) +
  labs(title = "Répartition du mix de production par région",
       x = NULL, y = "Part (%)", fill = "Filière") +
  theme_minimal()
```

```{r}
#| label: filiere-dominante-par-region

profil_region <- mix_region |> 
  group_by(region) |> 
  slice_max(part_pct, n = 1) |> 
  ungroup() |> 
  mutate(
    categorie = case_when(
      filiere == "nucleaire_mw" ~ "Nucléaire",
      filiere %in% c("eolien_mw", "solaire_mw", "hydraulique_mw", "bioenergies_mw") ~ "Renouvelable",
      filiere == "thermique_mw" ~ "Fossile",
      TRUE ~ "Autre"
    )
  )

profil_region
```

```{r}

```
```{r}

```

## Findings

<!-- TODO: what did you learn? This is your discussion: put your findings and
the original question back into perspective. Keep it honest: say what you
found, and what you did NOT find. -->

---

*This chapter is your semester's centrepiece. Revisit it as you learn new
techniques and make it your best work by the final gallery walk.*
