---
title: "Chapter 1 · My EDA project"
author: "Florian Escaffre"
date: today
---

> **How to fill this chapter in**
> This is your main portfolio piece. The numbered sections below follow the
> **IMRAD** structure of research: question ≈ *introduction*, getting the
> data + cleaning decisions ≈ *methods*, plotting the story ≈ *results*, and
> findings ≈ *discussion*. In the *Reproducible Workflows* course:
>
> - **Session 2** lab: find a dataset you care about on an open-data portal
>   (data.gouv.fr or your own country's), source it **reproducibly by code**,
>   run an EDA, clean the data, and commit it to git.
> - **Session 3** work block: finish and polish the chapter, pin your data and
>   environment, open a pull request for review, and publish the book to a URL.
>
> Everything between the `TODO` markers below is a template to replace with
> your own content.

## The question

<!-- TODO: In a sentence or two, what question are you trying to answer with this dataset?  -->

## Getting the data

The data is sourced **by code** from an open-data portal of your choice. Because
the download lives inside the document, it is recorded in the analysis itself;
anyone who re-renders this document gets the same data and the same results.
Record the dataset's **publisher and URL** in the text so the provenance is
explicit.

The only portal-specific part is the *fetch*. This template shows two ways;
use the one that fits your portal:

- **`{rdatagouv}`** (data.gouv.fr only, optional): search/download from R.
- **Any portal by URL**: `download.file()` works with any portal's direct
  download link.

```{r}
#| label: setup
#| message: false
# Load what you use. {rdatagouv} only if you're on data.gouv.fr and want it.
library(rdatagouv)    # search / download datasets from data.gouv.fr
library(janitor)        # clean column names, tabulate with tabyl()
library(naniar)         # visualise missing values
library(skimr)
library(dplyr) 
library(tidyverse)       # one-line data summary with skim()
```

```{r}
#| label: download
#| message: false
# TODO: replace with the code you wrote in the Session-2 lab.
# Option A — data.gouv.fr via {rdatagouv}:
# ds <- rdatagouv::get_dataset("5afbd35fb595086a31734624"); url <- ds$resources$url[1]
# Option B — any portal by direct URL:
# url <- "https://www.example-portal.org/.../my-data.csv"
# download.file(url, "data/my-data.csv")

ds <- readr::read_csv2("/home/ensai/Documents/portfolio-book-fe/portfolio-book/data/eco2mix-regional-cons-def.csv")
```

> **Optionally pin the fetched copy** with `{pins}` so the chapter reads from a
> versioned local board instead of hitting the network on every render:
>
> ```r
> board <- pins::board_folder("pins/", versioned = TRUE)
> pins::pin_write(board, dat, "my-data", type = "csv")
> dat    <- pins::pin_read(board, "my-data")
> ```


## A first look

```{r}
#| label: skim
#| fig-cap: "A one-line summary of the dataset."
# The skim() output gives you type, missingness and distributions at a glance.
ds |> 
  skimr::skim() 
```

```{r}
#| label: missingness
#| fig-cap: "Missing values across variables."
vis_miss(ds, "warn_large_data" = FALSE) #— where is data missing, and why might that be?
```

## Cleaning decisions {#sec-cleaning}

<!-- TODO: document the decisions you made while cleaning the data. What did you
rename, recode, or drop, and why? A defensible cleaning decision is a feature
of a good analysis, not an afterthought. -->

```{r}
# Clean names, drop empty column, fix column type
ds_clean <- ds |> 
  janitor::clean_names() |>
  janitor::remove_empty(which = "cols") |> 
  mutate(
    code_insee_region = as.integer(code_insee_region),
    
    across(
      .cols = c(
        eolien_mw, # Idéalement placé ici maintenant
        tco_thermique_percent, tch_thermique_percent, 
        tco_nucleaire_percent, tch_nucleaire_percent, 
        tco_eolien_percent, tch_eolien_percent, 
        tco_solaire_percent, tch_solaire_percent, 
        tco_hydraulique_percent, tch_hydraulique_percent, 
        tco_bioenergies_percent
      ),
      .fns = as.double
    )
  )

```
```{r}
#| label: skim
#| fig-cap: "A one-line summary of the cleaned dataset."
# The skim() output gives you type, missingness and distributions at a glance.
ds_clean |> 
  skimr::skim() 
```

Filtre pour garder seulement les données sur la Bretagne
```{r}
#| label: filtre-bretagne

ds_bzh <- ds_clean |> 
  filter(region == "Bretagne")

ds_bzh |> skimr::skim()
```

## Plotting the story

<!-- TODO: replace with your figures. Give every figure a caption and refer to
it in the text (e.g. "Fig. 1 shows that …"). If a figure uses randomness
(sampling, jitter, a simulation), fix the seed *inside* that cell with
`withr::with_seed(2025, { ... })` so the figure is identical on every render. -->

### Analyse de la bretagne
```{r}
#| label: fig-courbe-temporelle-conso-prod
#| fig-cap: "Courbe temporelle de la consommation en fonction de la production en bretagne"

ds_bzh |> 
  select(date_heure, consommation_mw, nucleaire_mw, eolien_mw, solaire_mw, 
         hydraulique_mw, thermique_mw, bioenergies_mw) |> 
  pivot_longer(-date_heure, names_to = "filiere", values_to = "mw") |> 
  ggplot(aes(x = date_heure, y = mw, color = filiere)) +
  geom_line(alpha = 0.7) +
  labs(title = "Consommation et production par filière — Bretagne",
       x = NULL, y = "MW", color = "Filière") +
  theme_minimal()
```

```{r}
#| label: saisonnalite-mensuelle

ds_bzh |> 
  mutate(mois = lubridate::month(date, label = TRUE)) |> 
  group_by(mois) |> 
  summarise(conso_moy = mean(consommation_mw, na.rm = TRUE), .groups = "drop") |> 
  ggplot(aes(x = mois, y = conso_moy)) +
  geom_col(fill = "steelblue") +
  labs(title = "Consommation moyenne par mois — Bretagne",
       x = NULL, y = "Consommation moyenne (MW)") +
  theme_minimal()
```

```{r}
#| label: mix-energetique-bzh

ds_bzh |> 
  select(date_heure, thermique_mw, eolien_mw, solaire_mw, hydraulique_mw, bioenergies_mw) |> 
  pivot_longer(-date_heure, names_to = "filiere", values_to = "mw") |> 
  group_by(filiere) |> 
  summarise(prod_totale = sum(mw, na.rm = TRUE), .groups = "drop") |> 
  mutate(part_pct = prod_totale / sum(prod_totale) * 100) |> 
  ggplot(aes(x = reorder(filiere, part_pct), y = part_pct)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "Répartition du mix de production — Bretagne (période complète)",
       x = NULL, y = "Part (%)") +
  theme_minimal()
```

```{r}
#| label: taux-couverture-bzh

ds_bzh |> 
  mutate(
    prod_totale = rowSums(across(c(thermique_mw, eolien_mw, solaire_mw, 
                                     hydraulique_mw, bioenergies_mw)), na.rm = TRUE),
    taux_couverture = prod_totale / consommation_mw * 100
  ) |> 
  summarise(
    taux_couverture_moy = mean(taux_couverture, na.rm = TRUE),
    taux_couverture_median = median(taux_couverture, na.rm = TRUE)
  )
```

## Findings

<!-- TODO: what did you learn? This is your discussion: put your findings and
the original question back into perspective. Keep it honest: say what you
found, and what you did NOT find. -->

---

*This chapter is your semester's centrepiece. Revisit it as you learn new
techniques and make it your best work by the final gallery walk.*
